---
title:  'Data Analysis Project: Knee Replacement Surgery'
subtitle: 'Analysis Proposal'
date: "November 10, 2016"
output:
  html_document:
    toc: yes
author:
- Gitika Jain (gitikaj2)
- Heather Sewak (hsewak2)
- Ionel Miu (imiu2)
- Jared Michael Coldert (jaredc2)
---
#  Knee Replacement Surgery Outcome

## Description
Knee Replacement Surgery uses implants made from metal and high density polyethylene to replace a diseased or worn out knee joint.  This dataset contains the details of a patient's knee replacement surgery.  The data includes the operating physician (converted to an integer value), anthropomorphic data about the patient, if computer navigation was utilized, the type of implants they received and two outcome metrics.  The first metric is comprised of several measurements indicating how well the surgeon believes the knee is functioning after surgery.  The second metric is a more subjective score measuring how well the patient is able to carry out daily activities.   

## Background information
This data has been collected from a Southern California Hospital in collaboration with the Shiley Center for Orthopedic Research.  The data is used for both clinical research and postoperative monitoring of implant survivorship.  All patient specific information has been removed from the data so it adheres to the Health Insurance Portability and Accountability Act (HIPAA).

## Statement
We hope to determine the impact a particular surgeon or the type of implants used has on the patient's outcome after surgery.  We are also interested in exploring the potential relationship between anthropomorphic factors and the patient's postoperative satisfaction.  Finally, we would like to understand if the considerable expense of computer navigation in knee replacement leads to better postoperative scores. Nevertheless, we keep our option opened to investigate other aspects that were not 'visibile' or never thought about it but might reveal themselfs as we move on with the project.

Here we may consider *KneeScore\_Surgeon* or *KennScore\_Patient* as response and *Age, Gendar, Weight, Height, BMI, Race, Side, Manufacturer* as predictors. *Gendar, Race, Side and Manufacturer* are some of the categorical predictors and *Age, Height, Weight and BMI* are numerical predictors.

## Load Data in R

```{r}
knee <- read.csv("knees.csv")

# dimensions of the dataset
dim(knee) 

# list the variables in dataset
names(knee)

# list the structure of dataset
str(knee) 

# list levels of factor Gender in dataset
levels(knee$Gender)

# print first 10 rows and 5 columns of dataset
head(knee[, 1: 5], n = 10)
```

###Knee Score Surgeon

```{r}
knee_model = lm(KneeScore_Surgeon ~ ., data = knee)
summary(knee_model)
```
**NOTE:** *knee_model* returns NA for some coefficients. That means that those variables are linearly related to others.

```{r}
knee_model_aic = step(knee_model, direction = "backward", trace = 0)
summary(knee_model_aic)
summary(knee_model_aic)$call
```
**NOTE:** *knee_model_aic* returns a model without NA coefficients. That's good as the algorithm was able to sense and eliminate those variables. 

```{r}
n = nrow(knee)
knee_model_bic = step(knee_model, direction = "backward", trace = 0, k = log(n))
summary(knee_model_bic)
summary(knee_model_bic)$call
```

**NOTE:** *knee_model_bic* returns a model without NA coefficients, as well.  
Also, it uses fewer variables when compared to *knee_model_aic*. 

```{r}
anova(knee_model_aic, knee_model)
```

**Note:** According to p-value which is big enough, null is not reject. That means that full model does not improve modeling
by adding extra variables.

```{r}
anova(knee_model_bic, knee_model)
```

**Note:** According to p-value, which is kind of small, the tendency is to reject the null. It all depends on the choise of level of confidence, alpha.


```{r}
anova(knee_model_bic, knee_model_aic)
```

**Note:** p-value is very small, therefore, null is rejected. Since these two models are nested (knee_model_bic is nested inside knee_model_aic) *knee_model_aic* offers improvments for the modeling.

```{r}
anova(knee_model_bic, knee_model_aic, knee_model)
```

**Note:** Again, modeling all three models at once, p-value is very small for *knee_model_aic*, therefore, null is rejected. Whereas *knee_model* does not reject the null.

**Comments:** Corelating all the above, so far, *knee_model_aic* seems to offer the best model.

```{r}
library(leaps)
all_knee_mod = summary(regsubsets(KneeScore_Surgeon ~ ., data = knee, really.big=T))
```















<!-- ```{r warning=FALSE, message=FALSE} -->
<!-- plot(fitted(knee_model_aic), resid(knee_model_aic), col = "magenta", pch = 20, cex = 1.5, xlab = "Fitted", ylab = "Residuals") -->
<!-- abline(h=0, lwd = 2, col = "olivedrab1") -->

<!-- qqnorm(resid(knee_model_aic), main = "Normal QQ Plot - Knee Model AIC", col = "purple4", pch = 20, cex = 1.5) -->
<!-- qqline(resid(knee_model_aic), col = "Orange", lwd = 3) -->

<!-- library(lmtest) -->
<!-- bptest(knee_model_aic) -->

<!-- shapiro.test(resid(knee_model_aic)) -->

<!-- ``` -->

<!-- ###Knee Score Patient -->
<!-- ```{r} -->
<!-- patient_model = lm(KneeScore_Patient ~ ., data = knee) -->
<!-- summary(patient_model) -->

<!-- patient_model_aic = step(patient_model, direction = "backward", trace = 0) -->
<!-- summary(patient_model_aic)$call -->

<!-- patient_model_bic = step(patient_model, direction = "backward", trace = 0) -->
<!-- summary(patient_model_bic)$call -->
<!-- ``` -->

<!-- ```{r} -->
<!-- anova(patient_model_aic, patient_model) -->
<!-- ``` -->

<!-- ```{r warning=FALSE, message=FALSE} -->
<!-- plot(fitted(patient_model_aic), resid(patient_model_aic), col = "green", pch = 20, cex = 1.5, xlab = "Fitted", ylab = "Residuals") -->
<!-- abline(h=0, lwd = 2, col = "black") -->

<!-- qqnorm(resid(patient_model_aic), main = "Normal QQ Plot - Patient Model AIC", col = "red", pch = 20, cex = 1.5) -->
<!-- qqline(resid(patient_model_aic), col = "navy", lwd = 3) -->

<!-- library(lmtest) -->
<!-- bptest(patient_model_aic) -->

<!-- shapiro.test(resid(patient_model_aic)) -->

<!-- ``` -->
